Inferring neural networks from fluorescent calcium imaging using partial correlation and sample weighting

### 2014-06-04 article.tex
Add todo
start to better explain directivity
Add references

% Introduce notation + goal in mathematical score
% TODO check convention for y_{ij}
One way to emphasize the connections between neurons is to represent them by a
(un)directed graph with $p$ nodes: each node represents a neuron, and each
edge $y_{ij}$ represents a direct link from neuron  $i$ to neuron $j$. If the
graph is undirected, then $y_{ij}$ is equal to $y_{ji}$  for all $i,j$.
The unsupervised task may be formulated as:
\textit{Given $X \in \mathcal{R}^{n \times p}$  a set of $p$ time series of
fluorescent calcium concentration of size $n$ from $p$ neurons, the goal is to
infer the connectivity graph $y \in \left\{0, 1\right\}^{p \times p}$.}
The main challenge in the inference of the network $y$
is to distinguish direct from indirect interactions [de la Fuente et al., 2004].
For instance with the simple interaction graph $A \rightarrow B \rightarrow C$,
we want to distinguish direct interactions $A \rightarrow B$ and $B \rightarrow C$
from the indirect interaction $A \rightarrow C$.

@@

% Correlation => Issue indirect effect
% Partial correlation => better estimate of the graph  => Symmetric measure
While the correlation coefficient is unable to distinguish between
direct and indirect interaction, the partial correlation coefficient
aims to infer only direct ones through the conditioning over all the other
variables. In previous studies [Shipley, 2002], it has been suggested
consider correlation coefficients of all possible orders [Shipley, 2002],
i.e. conditioned on every subsets of the set of $p-2$ other variables, in order
to take into account multiple indirect paths between a given pair of variables.
An obvious limitation to calculate high-order coefficient is the number of
samples in conditioned data sets \textcolor{red}{AJ: Est ce que ce ne sera pas
à la fois $p$ and $n$?}. Given the high number of samples $n$
in the Connectomics challenge, we only consider the highest order of partial
correlation coefficient.

Since the partial correlation is symmetric, it is not possible to infer the causal
mechanism behind the interaction (I directly causes J, J directly causes I or
both [de la Fuente, 2004]). Furthermore two other causal mechanisms may be implied
by a non-zero partial coefficient. First, the pair of variables may be induced
by a common (hidden) variable and second, I (respectively J) is conditionally
correlated to a hidden variable affecting J (resp. I) [de la Fuente, 2004].

% Introduce a custom solution to make the Y matrix asymmetric
% Introduce directivity
% TODO better link
%      add that we make the count matrix antisymmetric (count - count.T)
%      say that we normalize and stack
Some causal information (directivity of the links) were retrieved from the
data by comparing activity of each couple of neurons between two subsequent
time steps. The directivity method tries to detect variation of fluorescence
signal of a neuron $j$ due to a neuron $i$. This method counts the number of
times that $X_{t+1,j} - X_{t, i} \in [f_1, f_2]$ where $f_1$ and $f_2$ are
parameters of the method.

>>>>

% Introduce notation + goal in mathematical score
% TODO check convention for y_{ij}
One way to emphasize the connections between neurons is to represent them by a
(un)directed graph with $p$ nodes: each node represents a neuron, and each
edge $y_{ij}$ represents a direct link from neuron  $i$ to neuron $j$. If the
graph is undirected, then $y_{ij}$ is equal to $y_{ji}$  for all $i,j$.
The unsupervised task may be formulated as:
\textit{Given $X \in \mathcal{R}^{n \times p}$  a set of $p$ time series of
fluorescent calcium concentration of size $n$ from $p$ neurons, the goal is to
infer the connectivity graph $y \in \left\{0, 1\right\}^{p \times p}$.}
The main challenge in the inference of the network $y$
is to distinguish direct from indirect interactions \cite{de2004discovery}.
For instance with the simple interaction graph $A \rightarrow B \rightarrow C$,
we want to distinguish direct interactions $A \rightarrow B$ and $B \rightarrow C$
from the indirect interaction $A \rightarrow C$.

@@

% Correlation => Issue indirect effect
% Partial correlation => better estimate of the graph  => Symmetric measure
While the correlation coefficient is unable to distinguish between
direct and indirect interaction, the partial correlation coefficient
aims to infer only direct ones through the conditioning over all the other
variables. In previous study \cite{shipley2002cause}, it has been suggested
consider correlation coefficients of all possible orders,
i.e. conditioned on every subsets of the set of $p-2$ other variables, in order
to take into account multiple indirect paths between a given pair of variables.
An obvious limitation to calculate high-order coefficient is the number of
samples in conditioned data sets \textcolor{red}{AJ: Est ce que ce ne sera pas
à la fois $p$ and $n$?}. Given the high number of samples $n$
in the Connectomics challenge, we only consider the highest order of partial
correlation coefficient.

Since the partial correlation is symmetric, it is not possible to infer the causal
mechanism behind the interaction (I directly causes J, J directly causes I or
both ). Furthermore two other causal mechanisms may be implied
by a non-zero partial coefficient. First, the pair of variables may be induced
by a common (hidden) variable and second, I (respectively J) is conditionally
correlated to a hidden variable affecting J (resp. I) \cite{de2004discovery}.

% Introduce a custom solution to make the Y matrix asymmetric
% Introduce directivity
% TODO better link
%      add that we make the count matrix antisymmetric (count - count.T)
%      say that we normalize and stack
%      specify f_1 and f_2
In the hope to improve performance, some causal information (directivity of
the links) were added by stacking the partial correlation score matrix
with

from the
data by comparing activity of each couple of neurons between two subsequent
time steps. The directivity method tries to detect variation of fluorescence
signal of a neuron $j$ due to a neuron $i$. A connectivity score
between the two neurons $i$ and $j$ is given by
\[
z_{ij} = \sum_{t=1}^n I(X_{t+1,j} - X_{t, i} \in [f_1, f_2])
\]

where $f_1$ and $f_2$ are parameters of the method and $I(\bullet)$ is the
indicator function. Finally, a
